---
title: "Transcriptomics Tutorial - Count matrix preparation and analysis"
output:
  word_document:
    toc: false
  html_document:
    theme: readable
    highlight: default
    toc: false
    toc_float: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 2: Count matrix preparation and analysis

In this part of this tutorial, we will analyze a gene count matrix with 6 samples.

In addition to the sample preprocessed in part 1, the count matrix also contains 2 samples from dusk 
and 3 samples from dawn (the transition between nighttime and daylight).

- Gene Expression Omnibus (GEO) link to the dataset: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151567
- Published in: Zacharias AM, O'Connor CD, Topouza DG, Fang ZY, Ghazinejad H, Chen H, Duan Q, Ghasemlou N. Site- and cell-type-specific miRNA and mRNA genes and networks across the cortex, striatum, and hypothalamus. Commun Biol. 2025 Jul 1;8(1):969. doi: 10.1038/s42003-025-08371-7. PMID: 40594985; PMCID: PMC12218356.

In this tutorial we will:

1. Perform sample outlier detection using the gene count matrix.
2. Normalize gene counts.
3. Non-specific filtering of lowly variable genes.
3. Differential expression analysis using EdgeR.
4. Enrichment analysis of differentially expressed genes using g:Profiler.

> NOTE: So far in this class, we have had you run R code in an Rscript file (`*.R`).
> Another option is to run R code in an R Markdown file (`*.Rmd`), which allows you to combine R code with text and images.
> R Markdown uses markdown synatx to format text, similar to how Jupyter notebooks use markdown.
> R Markdown files can also be rendered into HTML, PDF, or Word documents, which is how this tutorial's files were created.
> Using either an Rscript or R Markdown file is fine. Do what is comfortable for you, but give both file-types a try.

This tutorial loads packages when they become relevant, rather than all at once.
However, **remember that it is best practice to load all packages at the beginning of your script**.

## 1. Setup

### 1.1 Project folder setup

We will remain in the same project folder as part 1 of this tutorial.
The following is how the directory structure should look at this point:

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
│   └── coldata.csv
│   └── counts.csv
│   └── id2name.csv
├── 1_qc/
├── 2_trim/
├── 3_align/
└── 4_quantify/
```

This time we will be using the `coldata.csv`, `counts.csv`, and `id2name.csv` files in the `0_data` directory, 
which contain the sample metadata, gene count matrix, and gene ID to gene name mapping, respectively.


### 1.2 Setup

Before we get started, let's setup our dataframes and orient ourselves.

Setup a folder to store data preparation scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 5_data_prep
cd 5_data_prep
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
├── 1_qc/
├── 2_trim/
├── 3_align/
├── 4_quantify/
└── 5_data_prep/
```

Create an R script or R markdown file in the `5_data_prep` directory.
Remember to take notes in our scripts about what the script is for, who made the script, when it was created, etc.
We will use this file to read in the count matrix, id2name, and coldata, and prepare output directories for the next steps.

``` {r eval=FALSE}
# Get your current working directory
getwd()

# If your working directory isn't the project folder, set it to the project folder.
setwd("/global/teaching-home/<your_username>/rnaseq_tutorial")

# Pathways -----------------------------------------------
## Inputs ============
data_dir <- "0_data"
counts_path <- file.path(data_dir, "counts.csv") # file.path is similar to `os.path.join` in Python.
coldata_path <- file.path(data_dir, "coldata.csv")
id2name_path <- file.path(data_dir, "id2name.csv")

## Outputs ===========
prep_dir <- "5_data_prep"
dataframes_dir <- file.path(prep_dir, "dataframes")
figures_dir <- file.path(prep_dir, "figures")
rdata_dir <- file.path(prep_dir, "rdata")

# Create the output directories using R instead of unix.
dir.create(dataframes_dir)
dir.create(figures_dir)
dir.create(rdata_dir)

# Load data -----------------------------------------------
counts <- read.csv(counts_path, row.names = 1)
coldata <- read.csv(coldata_path, row.names = 1)
id2name <- read.csv(id2name_path, row.names = 1)
```


Now that we've read in the dataframes, let's inspect them to make sure they look correct.

Beginning with the sample metadata (coldata).

``` {r eval=FALSE}
nrow(coldata) # number of samples
ncol(coldata) # number of columns in the coldata dataframe
head(coldata) # first 10 rows of the coldata dataframe
summary(coldata) # summary statistics of the coldata dataframe
```

Next, the count matrix

``` {r eval=FALSE}
nrow(counts) # number of samples
ncol(counts) # number of columns in the count matrix
head(counts) # first 10 rows of the count matrix
```

Next, the gene id to name mapping (id2name)

``` {r eval=FALSE}
nrow(id2name) # number of genes
head(id2name) # first 10 rows of the id2name dataframe
```

Do the gene names in the count matrix match the gene names in the id2name dataframe?

``` {r eval=FALSE}
# Are all gene
all(rownames(counts) %in% id2name$gene_name) # should return TRUE
all(id2name$gene_name %in% rownames(counts)) # should return TRUE
```

Finally, let's make sure that the order of samples is the same in both the coldata and count matrix dataframes.
The `Run` column in coldata corresponds to the fastq filenames, which are also now the column names of the count matrix.

``` {r eval=FALSE}
all(coldata$Run == colnames(counts)) # should return TRUE

# If the order of the samples is not the same, we can reorder the count matrix to match the coldata.
counts <- counts[, coldata$Run]
```

Let's pause to save our current progress.

``` {r eval=FALSE}
# Save the dataframes as CSV files
write.csv(counts, file.path(dataframes_dir, "raw_counts.csv"))
write.csv(id2name, file.path(dataframes_dir, "id2name.csv"))
write.csv(coldata, file.path(dataframes_dir, "coldata.csv"))
```

## 3 Sample outlier detection

Now that we've preliminarily wrangled our data and inspected it, let's perform sample outlier detection.
To do this, we will take a fairly straightforward approach of seeing how the samples cluster
together based on their gene expression profiles. 
If a sample is an outlier, we would expect it to cluster away from the other samples.

### 3.1 Variance stabilizing transformation

Before clustering, we need to apply variance stabilizing transformation (VST) to the count matrix.
VST is a transformation method implemented in the `DESeq2` package that prevents highly expressed genes
from dominating the clustering results. You can think of it as being similar to a log transformation, 
but it is more robust to the statistical properties of RNA-seq count data.

Prior to being able to do VST, we will

1. Create a DESeq2 dataset object from the count matrix and coldata.
2. Estimate the normalization size factors for the samples, allowing standard normalization of the count matrix.
3. Apply the VST to the normalized count matrix.

- DESeq2 vignette: https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
- DESeq2 paper: Love MI, Huber W, Anders S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol. 2014;15(12):550. doi: 10.1186/s13059-014-0550-8. PMID: 25516281; PMCID: PMC4302049.

``` {r eval=FALSE}
BiocManager::install("DESeq2") # if you haven't already installed DESeq2
library(DESeq2) # v1.48.2

dds <- DESeqDataSetFromMatrix(round(counts), coldata, design = ~ time)
dds <- estimateSizeFactors(dds)
vst <- varianceStabilizingTransformation(dds, blind = TRUE)
```

- `round(counts)` rounds the gene counts to the nearest integer, which is required for DESeq2.
- `design = ~ time` specifies that time is the experimental condition of interest for testing. This does not affect VST transformation.
- `blind = TRUE` means that the VST will be applied without using the sample information.

If you want, try to create a distribution plot (e.g., histogram) to compare the distribution of
gene counts before and after VST.

### 3.2 Clustering

We will now perform Principal Component Analysis (PCA) on the VST-transformed count matrix to visualize 
how the samples cluster together. We are using the `plotPCA` function from the `DESeq2` package.
You could also do t-SNE, hierarchical clustering, or other clustering methods, but PCA is a good starting point.

``` {r eval=FALSE}
# Perform principal component analysis
pca_data <- plotPCA(vst, intgroup = c("Run", "time"), returnData = TRUE)

# Calculate the percentage of variance explained by each principal component
percent_var <- round(100 * attr(pca_data, "percentVar"))
```

You could simply create a PCA plot with the `plotPCA` function, but we will take a more
advanced approach and use `ggplot2` to create a more customizable PCA plot.

``` {r eval=FALSE}
BiocManager::install("ggplot2") # if you haven't already installed ggplot2
library(ggplot2) # v4.0.1

# Colour points by the sample, indicated by `Run`
pca_plot_run <- ggplot(pca_data, aes(x = PC1, y = PC2, color = Run)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  theme_bw()
pca_plot_run # creates the plot in your Rstudio plots pane, or in the R markdown output document

# Colour points by the experimental condition, indicated by `time`
pca_plot_time <- ggplot(pca_data, aes(x = PC1, y = PC2, color = time)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  theme_bw()
pca_plot_time
```

Save the plots as PDF files

``` {r eval=FALSE}
ggsave(
  plot = pca_plot_run, path = figures_dir,
  filename = "pca_plot_run.pdf",
  width = 120, height = 90, units = "mm"
)
ggsave(
  plot = pca_plot_time, path = figures_dir,
  filename = "pca_plot_time.pdf",
  width = 120, height = 90, units = "mm"
)
```

Based on the PCA plots, do you see any samples that are outliers?
If you do, remove the outlier sample(s), re-run VST, re-run PCA, and re-plot the PCA plots.
Continue this process iteratively until you are satisfied that no outliers remain.

Example of removing an outlier sample:
``` {r eval=FALSE}
# Let's say the outlier sample is "SRRXXX"
counts_clean <- counts[, !colnames(counts) %in% c("SRRXXX")]
coldata_clean <- coldata[!coldata$Run %in% c("SRRXXX"), ]
```

In this case, we won't remove any samples, so can simply set `counts_clean` and `coldata_clean` to the original count matrix and coldata.

``` {r eval=FALSE}
counts_clean <- counts
coldata_clean <- coldata
```

At the end, we will have a 'clean' count matrix and coldata, which we will save as CSV files.

``` {r eval=FALSE}
write.csv(counts_clean, file.path(dataframes_dir, "clean_counts.csv"))
write.csv(coldata_clean, file.path(dataframes_dir, "clean_coldata.csv"))
```

## 4 Normalization

Although as an intermediate step, we already performed normalization of the count matrix,
we will perform differential expression analysis using EdgeR, which has its own normalization method.
In the world of RNA-seq analysis, it's good to know how both DESeq2 and EdgeR are widely used, so 
it's good to know how to use both packages.

- EdgeR manual: https://www.bioconductor.org/packages/devel/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf
- EdgeR normalization paper: Robinson MD, Oshlack A. A scaling normalization method for differential expression analysis of RNA-seq data. Genome Biol. 2010;11(3):R25. doi: 10.1186/gb-2010-11-3-r25. Epub 2010 Mar 2. PMID: 20196867; PMCID: PMC2864565.
- Recent EdgeR paper: Chen Y, Chen L, Lun ATL, Baldoni PL, Smyth GK. edgeR v4: powerful differential analysis of sequencing data with expanded functionality and improved support for small counts and larger datasets. Nucleic Acids Res. 2025 Jan 11;53(2):gkaf018. doi: 10.1093/nar/gkaf018. PMID: 39844453; PMCID: PMC11754124.

```{r eval=FALSE}
BiocManager::install("edgeR") # if you haven't already installed edgeR
library(edgeR) # v4.6.3

dge_list <- DGEList(counts_clean)
calc <- calcNormFactors(dge_list, method = "TMM")
counts_norm <- cpm(calc, normalized.lib.sizes = TRUE)
```

- `calcNormFactors` calculates normalization factors to scale the raw library sizes.
- `cpm` extracts the normalized counts from the DGEList object.

Try creating a plot to compare the distribution to compare gene counts before and after normalization.

## 5 Non-specific filtering of lowly variable genes

Now that our count matrix has been normalized, we will perform non-specific filtering of lowly variable genes.

To this we will:

1. Calculate the median absolute deviation (MAD) of each gene across all samples.
2. Determine the MAD threshold for filtering by calculating various percentiles of the MAD distribution.
3. Identify the genes that pass the MAD threshold.

### 5.1 Calculate the MAD of each gene

Beginning with calculating the MAD of each gene across all samples.

``` {r eval=FALSE}
# Rather than using a `for` loop, we can use the `apply` function to calculate the
#   MAD of each gene across all samples.
# `apply` is conceptually similar to list comprehension or the `map` function in Python.

# The resulting mads are saved as a one-column dataframe.
mads <- data.frame(mad = apply(X = counts_norm, MARGIN = 1, FUN = mad))
rownames(mads) <- rownames(counts_norm) # setting the rownames to the gene ids

# Inspect the resulting dataframe
head(mads)
```

Now let's identify an appropriate MAD threshold.
There is no one-size-fits-all threshold for filtering lowly variable genes.
We will use a data-driven approach to determine the threshold by
calculating various percentiles of the MAD distribution.

### 5.2 Determine the MAD threshold for filtering

The threshold we pick will be approximately where the MAD values go from 0 to non-zero values.

``` {r eval=FALSE}
quantile(mads$mad, probs = seq(0, 0.9, 0.01)) # 0th to 90th percentiles
```

At what percentile do the MAD values go from 0 to non-zero values?
In this case, it looks like the 49th percentile is a good threshold to use.
The MAD corresponding to the 49th percentile is 0.0007602902.
Therefore, we will remove all genes with a MAD < 0.0007602902.

### 5.3 Subset genes to keep based on the MAD threshold

``` {r eval=FALSE}
filt_threshold <- 0.0007602902
genes_to_keep <- rownames(mads)[mads$mad >= filt_threshold]
counts_filt <- counts_norm[genes_to_keep, ]
```

How many genes remain after filtering?

``` {r eval=FALSE}
cat("Number of genes before filtering: ", nrow(counts_norm), "\n")
cat("Number of genes after filtering: ", nrow(counts_filt), "\n")
```

### 5.4 Save the filtering results

We will save the dataframes as CSV files.
The variable `genes_to_keep` will be saved as an RDS file, which is a file format that can store R objects.

- RDS files can later be loaded into the R environment using the `readRDS` function.

``` {r eval=FALSE}
# Save the count matrix as a CSV file
write.csv(counts_filt, file.path(dataframes_dir, "counts_filt.csv"))

# Save the MAD values as a CSV file
write.csv(mads, file.path(dataframes_dir, "mads.csv"))

# Save the variables with genes to keep as an RDS file
saveRDS(genes_to_keep, file.path(rdata_dir, "genes_to_keep.rds"))
```

## 6 Differential expression analysis

Now that our data is prepared, let's perform differential expression analysis (DEA)!

Setup a folder to store differential expression analysis scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 6_dea
cd 6_dea
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
├── 1_qc/
├── 2_trim/
├── 3_align/
├── 4_quantify/
├── 5_data_prep/
└── 6_dea/
```

### 6.1 Setup

Let's load the normalized gene count matrix, coldata, id2name, and list of genes to keep.

Create an R script or R markdown file in the `6_dea` directory.
Remember to take notes in our scripts about what the script is for, who made the script, when it was created, etc.

``` {r eval=FALSE}
# Packages -----------------------------------------------
library(edgeR) # v4.6.3
library(dplyr) # v1.2.0
library(tibble) # v3.3.0
library(ggplot2) # v4.0.1

# Pathways -----------------------------------------------
## Inputs ===========
prep_dir <- "5_data_prep"
dataframes_dir <- file.path(prep_dir, "dataframes")
counts_path <- file.path(dataframes_dir, "clean_counts.csv")
id2name_path <- file.path(dataframes_dir, "id2name.csv")
coldata_path <- file.path(dataframes_dir, "clean_coldata.csv")

rdata_dir <- file.path(prep_dir, "rdata")
genes_to_keep_path <- file.path(rdata_dir, "genes_to_keep.rds")

## Outputs ===========
dea_dir <- "6_dea"
dataframes_dir <- file.path(dea_dir, "dataframes")
figures_dir <- file.path(dea_dir, "figures")
rdata_dir <- file.path(dea_dir, "rdata")

# Create the output directories using R instead of unix.
dir.create(dataframes_dir)
dir.create(figures_dir)
dir.create(rdata_dir)

# Load data -----------------------------------------------
counts <- read.csv(counts_path, row.names = 1)
coldata <- read.csv(coldata_path, row.names = 1)
id2name <- read.csv(id2name_path, row.names = 1)
genes_to_keep <- readRDS(genes_to_keep_path)
```

In order to perform DEA, we need to ensure that our experimental condition of interest is a `factor` variable.
We also want to make sure that the reference level of the factor is set to the control condition, which in this case is `dawn`.

``` {r eval=FALSE}
coldata$time # time before factorization
coldata$time <- factor(coldata$time, levels = c("dawn", "dusk"))
coldata$time # time after factorization
```

### 6.2 Prepare EdgeR objects

We will first create the DGEList object, which is the main data structure used by EdgeR.

``` {r eval=FALSE}
dge <- dge <- DGEList(counts = counts)
```

We will then create the design matrix, which specifies the experimental design and the comparisons we want to make.

- A conceptual guide to making design matrices: Law CW, Zeglinski K, Dong X, Alhamdoosh M, Smyth GK, Ritchie ME. A guide to creating design matrices for gene expression experiments. F1000Res. 2020 Dec 10;9:1444. doi: 10.12688/f1000research.27893.1. PMID: 33604029; PMCID: PMC7873980.
- Corresponding R vignette: https://www.bioconductor.org/packages//release/workflows/vignettes/RNAseq123/inst/doc/designmatrices.html

``` {r eval=FALSE}
design <- model.matrix(~ time, data = coldata)
head(design) # inspect the design matrix
```

- The formula `~ time` means that we will be contrasting dusk vs dawn, which are factors in the `time` column of the coldata dataframe.
- In the design matrix, each row is a sample.
- The first column of the design matrix is the y-intercept, which represents the baseline expression level of the control condition (dawn).
- The second column of the design matrix is the coefficient for the experimental condition (dusk), which represents the difference between dawn and dusk.

Save the design

``` {r eval=FALSE}
write.csv(design, file.path(dataframes_dir, "design.csv"))
```

### 6.3 EdgeR workflow

Starting with the DGEList that contains raw counts, we need to re-calculate normalization size factors.

``` {r eval=FALSE}
dge <- calcNormFactors(dge, method = "TMM")
```

Then estimate dispersion, which is a measure of the variability of gene expression across samples.
We perform this step because RNA-seq data has a statistical property called overdispersion, which means
that the variance of gene expression is greater than the mean.
We need to minimize this dispersion.

``` {r eval=FALSE}
dge <- estimateDisp(dge, design)
```

Filter out lowly variable genes using the list of genes to keep that we generated in part 5 of this tutorial.

``` {r eval=FALSE}
dge_filt <- dge[genes_to_keep, , ]
```

Fit the negative binomial generalized log-linear model to the filtered DGEList object.

``` {r eval=FALSE}
fit <- glmQLFit(dge_filt, design)
```

Perform the quasi-likelihood F-test to test for differential expression between the two conditions.

``` {r eval=FALSE}
qlf <- glmQLFTest(fit, coef = "timedusk")
```

### 6.4. Extract the results

We will get the results using edgeR's `topTags` function.

``` {r eval=FALSE}
results <- topTags(qlf, n = Inf, adjust.method = "BH", sort.by = "PValue")
results_df <- results$table
```

- `n = Inf`: means that we want to get all the results, not just the top N genes.
- `adjust.method = "BH"`: adjusting the P-values using the Benjamini-Hochberg (BH) method.
- `results$table`: contains the results as a data frame.

Merge the results with the id2name dataframe.

``` {r eval=FALSE}
results_id2name <- results_df |>
  tibble::rownames_to_column(var = "gene_name") |>
  dplyr::left_join(id2name, by = "gene_name")
```

- `tibble::rownames_to_column`: converts the rownames of the results dataframe to a column called `gene_name`.
- `dplyr::left_join`: merges the results dataframe with the id2name dataframe by the `gene_name` column.

How many genes are differentially expressed?

``` {r eval=FALSE}
sig_results <- results_id2name |>
  dplyr::filter(FDR < 0.05)
cat("\nNumber of genes tested: ", nrow(results_df), "\n")
cat("\nNumber of significant genes (FDR < 0.05): ", nrow(sig_results), "\n")
```

Save the results

``` {r eval=FALSE}
write.csv(results_id2name, file.path(dataframes_dir, "all_results.csv"))
write.csv(sig_results, file.path(dataframes_dir, "sig_results.csv"))
```

### 6.5 Plot the results

We will be visualizing the results using a volcano plot.
A volcano plot is a scatter plot that shows the relationship between the fold change and the statistical significance of each gene.
We will use the `ggplot2` package to create the volcano plot.

Add a column to the results dataframe that indicates whether each gene is significantly differentially expressed.

``` {r eval=FALSE}
plot_data <- results_id2name |>
  dplyr::mutate(significance = ifelse(FDR < 0.05, "Significant", "Not Significant"))
```

Get the BH signfiicance threshold

``` {r eval=FALSE}
bh_threshold <- max(sig_results$PValue)
```

Create the plot

``` {r eval=FALSE}
volcano_plot <- plot_data |>
  ggplot(aes(x = logFC, y = -log10(PValue), colour = significance)) +
  geom_point(alpha = 0.5) +
  scale_colour_manual(
    name = "Significance",
    values = c("Significant" = "red", "Not Significant" = "black")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_hline(yintercept = -log10(bh_threshold), linetype = "dotdash") +
  xlab("Log2 Fold Change") +
  ylab("-Log10 P-Value") +
  theme_bw()
volcano_plot
```

- `aes`: specifies the aesthetics of the plot, including the x and y axes and the color of the points.
- `geom_point`: creates the scatter plot.
- `scale_colour_manual`: sets the colors for the points based on significance.
- `geom_hline`: adds horizontal lines to indicate the significance thresholds.
- `xlab` and `ylab`: set the axis labels.
- `theme_bw`: applies a clean, white background theme to the plot.


Save the plot as a PNG file

``` {r eval=FALSE}
ggsave(
  plot = volcano_plot, path = figures_dir,
  filename = "volcano_plot.png",
  width = 180, height = 180, units = "mm"
)
```

### 6.6 Save the environment

We will save the R environment as an RData file, so we never have to run this testing again.

``` {r eval=FALSE}
save.image(file.path(rdata_dir, "dea.RData"))
```


## 7 Enrichment analysis

In the case of many signficant results, we can perform enrichment analysis to aid interpretation of results.

For this tutorial, we will use the `gprofiler2` package to perform enrichment analysis of the significant genes.

- The web-verison of g:Profiler can be found here: https://biit.cs.ut.ee/gprofiler/gost
- gprofiler2 vignette: https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
- gprofiler2 paper: Kolberg L, Raudvere U, Kuzmin I, Vilo J, Peterson H. gprofiler2 -- an R package for gene list functional enrichment analysis and namespace conversion toolset g:Profiler. F1000Res. 2020 Jul 15;9:ELIXIR-709. doi: 10.12688/f1000research.24956.2. PMID: 33564394; PMCID: PMC7859841.

Setup a folder to store enrichment analysis scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 7_enrichment
cd 7_enrichment
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
├── 1_qc/
├── 2_trim/
├── 3_align/
├── 4_quantify/
├── 5_data_prep/
├── 6_dea/
└── 7_enrichment/
```

### 7.1 Setup

Create an R script or R markdown file in the `7_enrichment` directory.
Remember to take notes in our scripts about what the script is for, who made the script, when it was created, etc.

Install the `gprofiler2` package if you haven't already.

``` {r eval=FALSE}
install.packages("gprofiler2")
library(gprofiler2) # 0.2.4
```

We will also be using `ggplot2` to visualize the enrichment results.

``` {r eval=FALSE}
library(ggplot2) # v4.0.1
```

Let's load the DEA results and create the output directories

``` {r eval=FALSE}
# Pathways -----------------------------------------------
## Input ===========
dea_dir <- "6_dea"
sig_results_path <- file.path(dea_dir, "dataframes", "sig_results.csv")

## Output ===========
enrich_dir <- "7_enrichment"
dataframes_dir <- file.path(enrich_dir, "dataframes")
figures_dir <- file.path(enrich_dir, "figures")
rdata_dir <- file.path(enrich_dir, "rdata")

dir.create(dataframes_dir)
dir.create(figures_dir)
dir.create(rdata_dir)

# Load data -----------------------------------------------
dea_res <- read.csv(sig_results_path, row.names = 1)
head(dea_res)
```

### 7.2 Perform enrichment analysis

We will perform enrichment analysis using the `gost` function from the `gprofiler2` package.

``` {r eval=FALSE}
gost_res <- gost(
  query = dea_res$gene_name,
  organism = "mmusculus",
  correction_method = "g_SCS",
  domain_scope = "annotated"
)
```

- The results list are stored in the gost_res variable.
- WThe g_SCS multiple testing correction method is a method developed by the creators of G:Profiler to account for many pathways being related to one another.
- The domain_scope = "annotated" argument means that only genes that are annotated to at least one pathway will be considered in the background set for this analysis. This limits the artiicial significance of results.

Let's also get the URL link to view the results in the g:Profiler web interface.

``` {r eval=FALSE}
gost_link <- gost(
  query = dea_res$gene_name,
  organism = "mmusculus",
  correction_method = "g_SCS",
  domain_scope = "annotated",
  as_short_link = TRUE
)
```

Get the results dataframe from the gost_res object.

``` {r eval=FALSE}
gost_df <- gost_res$result
res_df_chr <- apply(gost_df, 2, as.character) # convert all columns to character type, so we can save the dataframe
```

Save the results

``` {r eval=FALSE}
saveRDS(gost_res, file.path(rdata_dir, "gost_res.rds"))
write.csv(res_df_chr, file.path(dataframes_dir, "gost_results.csv"))
write.table(gost_link, file.path(dataframes_dir, "gost_link.txt"), row.names = FALSE, col.names = FALSE)
```

Plot the results as a bubble plot

``` {r eval=FALSE}
bubble_plot <- gost_df |>
  # subset(term_size > 100 & term_size < 500) |> # optionally subset by term size
  head(n = 10) |> #E get the top 10 results
  ggplot(aes(x = intersection_size / term_size, y = term_name)) +
  geom_point(aes(size = intersection_size, colour = -log10(p_value))) +
  xlab("overlap / term size") +
  ylab("term") +
  scale_colour_continuous(name = "-log10(adj. p)") +
  scale_size_continuous(name = "overlap") +
  theme_bw()
bubble_plot
```

- Although not always necessary, we can subset the results to only include terms/pathways of moderate size,
so that we exclude terms that are either too specific or too broad.

Save the plot as a PDF

``` {r eval=FALSE}
ggsave(
  plot = bubble_plot,
  path = figures_dir,
  filename = "bubble_plot.pdf",
  width = 90,
  height = 90,
  units = "mm"
)
```

Save the R environment

``` {r eval=FALSE}
save.image(file.path(rdata_dir, "enrichment.RData"))
```
