---
title: "Transcriptomics Tutorial - Raw sequencing reads to count matrix"
output:
  word_document:
    toc: false
  html_document:
    theme: readable
    highlight: default
    toc: false
    toc_float: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Raw sequencing reads to count matrix

In this part of this tutorial, we will perform quality checking, trimming, 
alignment, and quantification of one sample of RNA-seq data.

Sample "SRR11902288" is an mRNA-seq sample from the cerebral cortex of a naive/healthy male mouse, 
during dawn, which is the transition between night and daytime.

- Gene Expression Omnibus (GEO) link to the dataset: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151567
- Published in: Zacharias AM, O'Connor CD, Topouza DG, Fang ZY, Ghazinejad H, Chen H, Duan Q, Ghasemlou N. Site- and cell-type-specific miRNA and mRNA genes and networks across the cortex, striatum, and hypothalamus. Commun Biol. 2025 Jul 1;8(1):969. doi: 10.1038/s42003-025-08371-7. PMID: 40594985; PMCID: PMC12218356.

Given time constraints, we will only process this one sample.
The FastQC files have also been randomly subset to only contain 1 million reads, instead of 50 million reads.
The same steps can be applied to all samples in a dataset. Specifically, we will:

1. Perform quality control of raw sequencing reads using FastQC and MultiQC.
2. Trim low-quality bases and adapter sequences using fastp.
3. Align trimmed reads to the mouse reference genome using Hisat2.
4. Quantify gene expression using StringTie.

> NOTE: Refrain from copying and pasting code because it can introduce hidden characters and formatting errors.
> Remember at each step to check the .out and .err files before moving on to the next step.

## 1. Setup

### 1.1 Project folder setup

Make a directory called `rnaseq_tutorial` and change into it.

``` {bash eval=FALSE}
# "~" is equivalent to `/global/teaching-home/<your_username>`
mkdir ~/rnaseq_tutorial
cd ~/rnaseq_tutorial
```

Use `scp` to securly copy the tutorial files from the course directory to your project folder.

``` {bash eval=FALSE}
# "~" is equivalent to `/global/teaching-home/<your_username>`
scp -r /global/teaching-project/sg3240000/rnaseq_tutorial/* .
```

Check that the files have been successfully copied.

``` {bash eval=FALSE}
cd ~/rnaseq_tutorial
ls
```

You should see the following directory called `0_data` was added to your working directory.
This folder contains the raw sequencing reads in FASTQ format, stored in a directory called `raw_reads`.
This folder also contains the reference genome and annotation files, stored in a directory called `gencode`.

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   │   ├── SRR11902288_1.fastq.gz
│   │   └── SRR11902288_2.fastq.gz
│   ├── gencode/
│   │   ├── gencode.vM32.primary_assembly.annotation.gtf
│   │   └── GRCm39.primary_assembly.genome.fa
│   └── coldata.csv # used for tutorial p2.
|   └── counts.csv  # used for tutorial p2.
│   └── id2name.csv # used for tutorial p2.
```

### 1.2 Inspect the FASTQ files

Use `zless` inspect the contents of these FASTQ files (https://en.wikipedia.org/wiki/FASTQ_format).
`zless` Allows you to interactively scroll up and down a compressed file. To exit, press `q` on your keyboard.

``` {bash eval=FALSE}
zless 0_data/raw_reads/SRR11902288_2.fastq.gz
q
zless 0_data/raw_reads/SRR11902288_1.fastq.gz
q
```

### 1.3 Submitting jobs to the SLURM scheduler

In this section, we will prepare a template bash script for submitting jobs to the CAC, 
and remind you how to submit a job. For more info see: 
https://jstaf.github.io/hpc-intro/scheduler/ and https://slurm.schedmd.com/documentation.html

``` {bash eval=FALSE}
#!/bin/bash
#SBATCH --job-name=job_name
#SBATCH --mail-type=ALL
#SBATCH --mail-user=YourEmail
#SBATCH --cpus-per-task=1
#SBATCH --mem=10MB
#SBATCH --time=0-0:10:00  # Day-Hours-Minutes-Seconds
#SBATCH --output=job_name.out
#SBATCH --error=job_name.err
# Title: Example bash script
# Author: <Your name>
# Date: <Date>
# Email: <Your email>
# Notes:
#

echo Job started at "$(date +%T)"

<Your Job Goes here>

echo Job ended at "$(date +%T)"
```

Submit your job.

``` {bash eval=FALSE}
sbatch example.sh
```

To check status of job(s) submitted:

``` {bash eval=FALSE}
squeue -u <your username>
# or
watch squeue -u <your username>
```

To cancel a job:

``` {bash eval=FALSE}
scancel <job id number>
# or
scancel -n <job name>
# or
scancel -u <your username> # use with caution!
```

### 1.4 Parallelization within a script

In this tutorial and in this assignment you will have to process multiple files in the same way.
To do this efficiently, you could create a bash script for each file and submit multiple jobs at the same time.
In this tutorial, we will create a function that processes a single file, and apply it to all files with parallelization.

- Parallel documentation: https://linux.die.net/man/1/parallel

Example of a bash script with parallelization:

```
#!/bin/bash
#SBATCH --job-name=job_name
#SBATCH --mail-type=ALL
#SBATCH --mail-user=YourEmail
#SBATCH --cpus-per-task=2
#SBATCH --mem=10MB
#SBATCH --time=0-0:10:00  # Day-Hours-Minutes-Seconds
#SBATCH --output=job_name.out
#SBATCH --error=job_name.err
# Title: Example bash script
# Author: <Your name>
# Date: <Date>
# Email: <Your email>
# Notes:
#

echo Job started at "$(date +%T)"

# Dependencies

# Commands to run your analysis
# Function
echo_text() {
    echo "Hello $1"
}

# Execute
export -f echo_text
parallel --jobs 2 echo_text ::: Alice Bob Charlie

echo Job ended at "$(date +%T)"
```

Submit your job.

``` {bash eval=FALSE}
sbatch example.sh
```

## 2 Quality control of raw sequencing reads

Before we can analyze the sequencing reads, we have to check their quality and potentially clean them.

FastQC is a quality control tool that generates a report for each FASTQ file. 
MultiQC aggregates the FastQC reports into a single report.
From the FastQC and MultiQC reports, we can see whether cleaning of reads is necessary.
In this case, we will use fastp to trim low-quality bases and adapter sequences from the reads.
Although not in this tutorial, we would also typically run FastQC and MultiQC on the trimmed reads.

Setup a folder to store quality control scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 1_qc
cd 1_qc
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   ├── gencode/
└── 1_qc/
```

### 2.1 Run FastQC

The following script will run FastQC on all FASTQ files in the `0_data/raw_reads` directory, 
and then run MultiQC to generate a single report. Specifically, the script will:

1. Create a directory called `fastqc_reports` to store the FastQC reports.
2. Load the FastQC module.
3. Run FastQC on all FASTQ files in the `0_data/raw_reads` directory with parallelization.

- FastQC documentation: https://www.bioinformatics.babraham.ac.uk/projects/fastqc/

Insert the following into a bash script, replacing `<Your Job Goes here>`.

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 fastqc/0.12.1

# Variables
READSPATH=../0_data/raw_reads # path to the directory containing the raw FASTQ files
OUTDIR=./fastqc_out  # path to the output directory where FastQC reports will be saved
mkdir -p "$OUTDIR"        # create the output directory if it does not already exist
MAX_JOBS=2                # maximum number of parallel processes

# Function
run_fastqc() {
    file="$1"
    echo "Running FastQC on $file"
    fastqc -o "$OUTDIR" "$file"
}

# Execute
export -f run_fastqc
export OUTDIR

find "$READSPATH" -name "*.fastq.gz" | \
    parallel --jobs "$MAX_JOBS" run_fastqc {}
```

Recommended 2 CPUS, 1GB of memory, and 10 minutes of time.

Notes:

- `mkdir -p "$OUTDIR"`: creates the output directory if it does not already exist.
- `READSPATH`: path to the directory containing the raw FASTQ files.
- `OUTDIR`: path to the output directory where FastQC reports will be saved. `mkdir -p` creates it if it does not already exist.
- `MAX_JOBS`: the number of FastQC processes to run at the same time. Set to match `--cpus-per-task`.
- `run_fastqc()`: a Bash function that runs FastQC on a single file passed as `$1`.
- `export -f run_fastqc`: makes the function available to the subshells that `parallel` spawns. Without this, parallel workers would not recognize `run_fastqc` as a command. `-f` specifies that the name refers to a function, not a variable.
- `export OUTDIR`: makes the `OUTDIR` variable available inside each worker subshell so `run_fastqc` can use it.
- `find "$READSPATH" -name "*.fastq.gz"`: generates the list of input files to process.
- `| parallel --jobs "$MAX_JOBS" run_fastqc {}`: pipes each file path to `parallel`, which calls `run_fastqc` with that path (`{}`) as the argument, running up to `MAX_JOBS` calls concurrently.  
- `\`: backslash at the end of a line allows the command to continue on the next line for readability. Make sure there are no spaces or text after the `\`.

### 2.2 Run MultiQC

MultiQC aggregates the FastQC reports into a single report.

- MultiQC documentation: https://docs.seqera.io/multiqc/
- MultiQC paper: Ewels P, Magnusson M, Lundin S, Käller M. MultiQC: summarize analysis results for multiple tools and samples in a single report. Bioinformatics. 2016 Oct 1;32(19):3047-8. doi: 10.1093/bioinformatics/btw354. Epub 2016 Jun 16. PMID: 27312411; PMCID: PMC5039924.

#### 2.2.1 Install MultiQC

MultiQC is built in python and can be installed using pip in a virtual environment.
Conda also works, but is not supported by the CAC.

Set your working directory to the project folder & run the following code in your terminal.

``` {bash eval=FALSE}
module load StdEnv/2020 python/3.9.6
mkdir -p env
python3 -m venv env/multiqc_env     # create the virtual environment
source env/multiqc_env/bin/activate # activate the virtual environment
pip install "multiqc==1.14"         # takes approx 3 minutes to install
deactivate                          # deactivate the virtual environment
```

#### 2.2.2 Run MultiQC

Navigate back to the `1_qc` directory and create a bash script to run MultiQC.

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 python/3.9.6
source ../env/multiqc_env/bin/activate # activate the virtual environment w/ multiqc

# Execute multiqc
multiqc \
    --outdir multiqc_out \
    --cl-config "fastqc_config: { fastqc_theoretical_gc: mm10_txome }" \
    fastqc_out
deactivate
```

Recommended 1 CPU, 20 MB, and 10 minutes of time.

Notes:

- `--outdir`: path to the output directory where the MultiQC report will be saved.
- `--cl-config`: the theoretical GC content of the mouse transcriptome (mm10_txome).
- `fastqc_out`: path to the directory containing the FastQC reports.

#### 2.2.3 Inspect the MultiQC report

Open the multiqc_report.html file in your browser to inspect the MultiQC report.

## 3. Trim low-quality bases and adapter sequences

Fastp is a tool that trims low-quality bases and adapter sequences from raw sequencing reads.

- Fastp documentation: https://github.com/opengene/fastp
- Fastp paper: Chen S. fastp 1.0: An ultra-fast all-round tool for FASTQ data quality control and preprocessing. Imeta. 2025 Sep 9;4(5):e70078. doi: 10.1002/imt2.70078. PMID: 41112039; PMCID: PMC12527978.

Setup a folder to store trimming scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 2_trim
cd 2_trim
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
├── 1_qc/
└── 2_trim/
```

### 3.1 Run fastp

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 fastp/1.0.1

# Variables
READSPATH=../0_data/raw_reads # path to the directory containing the raw FASTQ files
TRIMREADSDIR=./trimmed_reads  # path to the output directory w/ trimmed FASTQ files
REPORTSDIR=./fastp_reports    # path to the directory w/ fastp reports
mkdir -p "$TRIMREADSDIR" "$REPORTSDIR"
MAX_JOBS=2

# Function
run_fastp() {
    file="$1"
    base=$(basename "$file" _1.fastq.gz)
    echo "Running Fastp on $file"
    fastp \
        -q 20 \
        -l 30 \
        -i "$file" \
        -I "${READSPATH}/${base}_2.fastq.gz" \
        -o "${TRIMREADSDIR}/${base}_1.trim.fastq.gz" \
        -O "${TRIMREADSDIR}/${base}_2.trim.fastq.gz" \
        -h "${REPORTSDIR}/${base}_fastp.html" \
        -j "${REPORTSDIR}/${base}_fastp.json"
}

# Execute
export -f run_fastp
export READSPATH TRIMREADSDIR REPORTSDIR

find "$READSPATH" -name "*_1.fastq.gz" | \
    parallel --jobs "$MAX_JOBS" run_fastp {}
```

Recommended 2 CPUS, 7GB, 10 minutes of time.

Notes:

- `base=$(basename "$file" _1.fastq.gz)` extracts the sample name from the input file path, removing the `_1.fastq.gz` suffix.
- `-q`: phred quality score threshold for trimming low-quality bases.
- `-l`: minimum length of reads to keep after trimming.
- `-i`: path to the input FASTQ file for read 1.
- `-I`: path to the input FASTQ file for read 2.
- `-o`: path to the output FASTQ file for trimmed read 1.
- `-O`: path to the output FASTQ file for trimmed read 2.
- `-h`: path to the HTML report file.
- `-j`: path to the JSON report file.

Check out the fastp report to inspect the results of trimming. 
Optionally you could re-run FastQC and MultiQC on the trimmed reads.

If any samples were to continue having low quality reads, we would remove that samples from analysis.

## 4. Alignment

Once sequencing reads are cleaned and high quality, we can align them to the reference genome. 
In this tutorial, we will use Hisat2 to align the reads to the mouse reference genome.

- Hisat2 manual: https://daehwankimlab.github.io/hisat2/manual/
- Hisat2 paper: Kim D, Paggi JM, Park C, Bennett C, Salzberg SL. Graph-based genome alignment and genotyping with HISAT2 and HISAT-genotype. Nat Biotechnol. 2019 Aug;37(8):907-915. doi: 10.1038/s41587-019-0201-4. Epub 2019 Aug 2. PMID: 31375807; PMCID: PMC7605509.

Setup a folder to store alignment scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 3_align
cd 3_align
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
├── 1_qc/
├── 2_trim/
└── 3_align/
```

## 4.1 Build Hisat2 index

We will align the sequencing reads to the mouse genome (DNA sequence) and annotation files 
from the GENCODE project (https://www.gencodegenes.org/mouse/).
Before we can do this, we need to **index** the reference genome, so that Hisat2 can efficiently search it 
for matches to the sequencing reads.

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 hisat2/2.2.1

# Variables
GENOMEDIR=../0_data/gencode # path to the reference genome files
INDEXDIR=./hisat2_index     # path to the output directory within the 3 _align dir
mkdir -p "$INDEXDIR"        # create the output directory if it does not already exist

# Body
echo "Extracting splice sites and exons from the GTF file..."
hisat2_extract_splice_sites.py ${GENOMEDIR}/gencode.vM32.primary_assembly.annotation.gtf > ${INDEXDIR}/GRCm39.ss
hisat2_extract_exons.py ${GENOMEDIR}/gencode.vM32.primary_assembly.annotation.gtf > ${INDEXDIR}/GRCm39.exon

echo "Building Hisat2 index..."
hisat2-build \
    -p 20 \
    -f ${GENOMEDIR}/GRCm39.primary_assembly.genome.fa \
    --ss ${INDEXDIR}/GRCm39.ss \
    --exon ${INDEXDIR}/GRCm39.exon \
    ${INDEXDIR}/GRCm39_index
```

Recommended 10 CPUS, 150 GB, and 2 hours of time. It will probably take more like an hour, but we're requesting more time in case.

- `-p`: number of threads to use for indexing. Set to match `--cpus-per-task` for the SLURM scheduler.
- `-f`: path to the input genome FASTA file.
- `--ss`: path to the splice site file generated from the GTF annotation file.
- `--exon`: path to the exon file generated from the GTF annotation file.
- `INDEXDIR`: path to the output directory for the Hisat2 index.

## 4.2 Align reads with Hisat2

Now that the reference genome is indexed, we can align the trimmed reads to it using Hisat2.
There's just one sample in this tutorial, so no paralleization is needed.

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 hisat2/2.2.1

# Variables
SAMDIR=./hisat2_sam        # path to the output directory with SAM files
REPORTDIR=./hisat2_reports # path to the output directory with Hisat2 reports
mkdir -p "$SAMDIR" "$REPORTDIR"

# Alignment
hisat2 -p 24 \
    -x "./hisat2_index/GRCm39_index" \
    -1 "../2_trim/trimmed_reads/SRR11902288_1.trim.fastq.gz" \
    -2 "../2_trim/trimmed_reads/SRR11902288_2.trim.fastq.gz" \
    --dta \
    --time \
    --verbose \
    --summary-file "${REPORTDIR}/SRR11902288_summary.txt" \
    -S "${SAMDIR}/SRR11902288.sam"
```

Recommended 24 CPUS, 10 GB, and 10 minutes of time.

Notes:

- `-p`: number of threads to use for alignment. Set to match `--cpus-per-task` for the SLURM scheduler.
- `-x`: path to the Hisat2 index prefix.
- `-1`: path to the input FASTQ file for read 1.
- `-2`: path to the input FASTQ file for read 2.
- `--dta`: stands for `downstream-transcriptome-assembly`. Tailors alignments for transcript assemblers like StringTie by requiring longer anchor lengths for de novo discovery of splice sites.
- `--time`: print the time to load the index files and align the reads.
- `--verbose`: print verbose output, useful for debugging.
- `--summary-file`: path to the output summary file with alignment statistics.
- `-S`: path to the output SAM file with aligned reads.

There is no strict threshold, but if any sample has a clearly low alignment rate  compared to the 
other samples, we would remove that sample from downstream processing and analysis.
A good default threshold may be an 80% unique and overall alignment rate, but can certainly vary across studies.

## 4.3 Convert SAM to BAM, sort, and mark duplicates

After aligning the reads, we need to convert the SAM file to a BAM file, sort it, and mark duplicates.
We convert the files from SAM to BAM so that they take up less space and can be processed more efficiently.
We sort the BAM file by genomic coordinates so it can be analyzed more efficiently by downstream tools.
Finally, we mark duplicate reads that may have been introduced during PCR amplification. 
This doesn't affect quantification, but could be important if we ever want to call variants in these data.

- Samtools manual: http://www.htslib.org/doc/samtools.html 
- Samtools duplicate marking documentation: http://www.htslib.org/algorithms/duplicate.html

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 samtools/1.18

# Variables
SAMDIR=./hisat2_sam
BAMDIR=./hisat2_bam
mkdir -p "$BAMDIR"

# Code
samtools view -@ 4 -b "${SAMDIR}/SRR11902288.sam" | \
    samtools collate -@ 4 -u -O - | \
    samtools fixmate -@ 2 -m -u - - | \
    samtools sort -@ 12 -l 0 - | \
    samtools markdup -@ 2 - "${BAMDIR}/SRR11902288.sort.markdup.bam"
samtools index "${BAMDIR}/SRR11902288.sort.markdup.bam"
```

Recommended 24 CPUS, 10 GB, and 20 minutes of time.

Notes:

- You can play around with the number of threads for each command if you want.
- `|` pipes the output of one command to the input of the next command, without having to create intermediate variables.
- `samtools view -@ 4 -b`: converts the SAM file to BAM format using 4 threads.
    - `-b`: output BAM format.
- `samtools collate -@ 4 -u -O -`: groups reads by name
    - `-u`: output uncompressed BAM.
    - `-O`: output to the `stdout`, which is the standard output environment. Allows the output to be piped.
- `samtools fixmate -@ 2 -m -u - -`: fixes read mate information using 2 threads. Mates are reads that correspond to each other in paired-end sequencing.
    - `-u`: output uncompressed BAM.
    - `-m`: adds a score for the alignment quality of both mates in a read pair.
- `samtools sort -@ 12 -l 0 -`: sorts the BAM file by genomic coordinates using 12 threads.
    - `-l 0`: output uncompressed BAM.
- `samtools markdup -@ 2 -`: marks duplicate reads using 2 threads.
- `samtools index`: creates an index file for the sorted BAM file, which speeds up downstream analysis.

Now that your BAM files have been created, you could remove the SAM files to save space.

## 5. Quantification

Now that reads are aligned to the reference genome, we can quantify gene expression using StringTie.
First, StringTie is first run on each sample to estimate transcript abundances, resulting in a GTF file for each sample.
Then, we combine all GTFs into a single count matrix using the `tximport` R package.

- StringTie manual: https://ccb.jhu.edu/software/stringtie/index.shtml?t=manual
- StringTie paper: Pertea M, Pertea GM, Antonescu CM, Chang TC, Mendell JT, Salzberg SL. StringTie enables improved reconstruction of a transcriptome from RNA-seq reads. Nat Biotechnol. 2015 Mar;33(3):290-5. doi: 10.1038/nbt.3122. Epub 2015 Feb 18. PMID: 25690850; PMCID: PMC4643835.

For this tutorial, we only 1) quantify gene expression and 2) combine counts across samples.
As you will see in the StringTie manual, quantification with StringTie is usually a four-step process.

1. Assemble transcripts (known and novel).
2. Merge transcript assemblies from all samples into a single GTF file.
3. Re-estimate transcript abundances using the merged GTF file.
4. Combine counts across samples into a single count matrix.

The tutorial is skipping the transcript assembly and merging steps; however, 
you **will** need to do these steps in the assignment.
It will be up to you to figure out exactly how to do this by looking at the
StringTie manual and/or online resources.

Setup a folder to store quantification scripts and results. Enter the new directory.

``` {bash eval=FALSE}
# Paths assume that your current working directory is the project folder.
mkdir 4_quantify
cd 4_quantify
```

```
rnaseq_tutorial
├── 0_data/
│   ├── raw_reads/
│   └── gencode/
├── 1_qc/
├── 2_trim/
├── 3_align/
└── 4_quantify/
```

### 5.1 Run StringTie

StringTie results in a GTF with the transcript abundance results.
In this tutorial, StringTie is only run once because there is only one sample. 
In practice, StringTie would be run for each sample in the dataset.

``` {bash eval=FALSE}
# Dependencies
module load StdEnv/2023 stringtie/3.0.1
# Variables
GTFDIR=./gtfs
mkdir -p ${GTFDIR}/SRR11902288

# Body
stringtie -p 12 \
    -G ../0_data/gencode/gencode.vM32.primary_assembly.annotation.gtf \
    -e \
    -B \
    -o ${GTFDIR}/SRR11902288/SRR11902288.gtf \
    ../3_align/hisat2_bam/SRR11902288.sort.markdup.bam
```

Recomended 12 CPUS, 600 MB, and 10 minutes of time.

Notes:

- `-p`: number of threads to use for quantification.
- `-G`: path to the GTF annotation file.
- `-e`: only estimate the abundance of the given reference transcripts, without assembling novel transcripts.
- `-B`: outputs additional quantified feature count files.
- `-o`: path to the output GTF file with transcript abundance results.

Look at the first 10 lines of the GTF file to inspect the results.

```{bash eval=FALSE}
head gtfs/SRR11902288/SRR11902288.gtf
```

### 5.2 Combine counts across samples

We will combine the gene counts across samples into a single count matrix using the R package tximport.

Your working directory for Rscripts should be the project folder (`rnaseq_tutorial`).

In **R**, create a folder to store the count matrix and gene ID to gene name mapping files.

```{r eval=FALSE}
dir.create(file.path("4_quantify", "counts")) # file.path is similar to `os.path.join` in Python.
```

### 5.2.1 Install the dependencies

```{r eval=FALSE}
# Install BiocManager if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
}
BiocManager::install("tximport")
```

Load the R package.

```{r eval=FALSE}
library(tximport)
```

Write a comment next to the code indicating what version of tximport you are using.

You will also need to install and load the `dplyr` R package for data wrangling.

```{r eval=FALSE}
BiocManager::install("dplyr")
library(dplyr)
```

### 5.2.2 List the count files

```{r eval=FALSE}
files <- list.files(
  file.path("4_quantify", "gtfs"),
  full.names = TRUE, # return the entire path
  recursive = TRUE,  # search within folders
  pattern = "t_data.ctab$" # find all files that end with "t_data.ctab"
)
# convert the vector of paths into a named vector, where the names are the sample names
names(files) <- basename(dirname(files))
```

### 5.2.3 Map gene/transcript ids to gene names

Tximport expects a mapping of transcript IDs to gene names, which can be extracted from one of the ctab files.

``` {r eval=FALSE}
# Read in one of the files
tmp <- read.delim(files[1])

# Extract the transcript id and gene name information
tx2gene <- tmp[, c("t_name", "gene_name")]
```

Separately, we want to know the mapping between gene ids and gene names.

``` {r eval=FALSE}
id2name <- tmp[, c("gene_id", "gene_name")] |>
  subset(! gene_name == ".") |>
  distinct()
```

- `subset(! gene_name == ".")` removes rows where the gene name is missing (represented by a dot).
- `distinct()` removes duplicate rows, so that each gene ID is only represented once.

`|>` is the pipe operator, which passes the output of one function to the next function.
In this case, we're passing the `counts` dataframe to the `subset` function, which will remove features
with an unknown gene name, indicated by "."
The `%>%` operator is also a pipe operator that is part of the `dplyr` package.
They have subtle differences but are overall similar, so we'd recommend using `|>` 
because it's built into base R.

### 5.2.4 Combine counts across samples

We will use the tximport function to combine the counts.

- tximport documentation: https://www.bioconductor.org/packages//release/bioc/vignettes/tximport/inst/doc/tximport.html#StringTie

``` {r eval=FALSE}
txi <- tximport(files, type = "stringtie", tx2gene = tx2gene, readLength = 50)
gene_counts <- txi$counts

# Again removing genes lacking a known gene name for simplicity
gene_counts <- gene_counts[! rownames(gene_counts) %in% ".", ]
```

### 5.2.5 Save the count matrix and id2name

Finally, we will save results to CSV files.

``` {r eval=FALSE}
write.csv(gene_counts, file = file.path("4_quantify", "counts", "counts.csv"))
write.csv(id2name, file = file.path("4_quantify", "counts", "id2name.csv"))
```
